### 2021-06-13 Manuscript.Rmd
updated manuscript

Siphonophores are highly complex colonial organisms, consisting of bodies (called zooids) that share a common gastrovascular cavity but that are functionally specialized to specific tasks, including feeding, swimming, and reproducing. In this paper, we use RNA-seq to investigate the functional specialization of siphonophore zooids at a molecular level. We investigated expression patterns in five siphonophore zooids, and one specialized tissue (the pneumatophore, a gas filled float), across seven siphonophore species, in addition to unique species-specific zooids, and where possible developing and mature zooids. We used classical differential gene expression analyses within species to identify patterns that define particular zooid types, and gain insight into the genes that may be playing a role in the structure, maintenance, and functioning of zooids. We identified hundreds of high transcript abundance genes that are specific to particular zooids, as well as a number of putative transcription factors that have higher expression in particular zooids and developmental stages. We also examined whether gene expression patterns could distinguish novel, distinct species-specific zooid types within three species. We found clear evidence for morphological and functional differences among the tentacular palpon and gastrozooid of _Physalia physalis_, including putative toxin genes with distinct expression profiles.  By contrast we found few morphological or functional differences between yellow and white gastrozooids in _Bargmannia elongata_, or between gastric and B-palpons in _Agalma elegans_. Lastly, we compare gene expression across species, to identify putative clade/lineage or zooid specific expression patterns. We use TPM10K expression values, a metric which normalizes TPM to account for different sequencing depths among species, as well as ratios of expression counts to control for unknown species and gene-specific counting-efficiency coefficients. Using classical analyses using strict orthologs (Strict Ortholog Filtering of Trees), we identified 494 species-variable genes, and 349 tissue-variable genes. We also introduce new phylogenetically informed methods that involve filtering expression data using branches in gene trees that are equivalent to branches in the species tree (XXXXX name STEEL). We find that gene expression patterns tend to be largely consistent among species, as most expression ratios show very little change across species tree equivalent gene tree branches. We also identified a number of branches with very large expression changes, representing putative linage specific shifts in expression.

@@

Colonial animals, consisting of genetically identical bodies that are physically and physiologically connected, can be found across the metazoan tree of life [@hiebert2020coloniality]. Functional specialization of such bodies has evolved multiple times within colonial animals, with siphonophores in particular showing the greatest diversity of functionally specialized bodies [@Beklemishev1969]. Siphonophores are highly complex, colonial “superorganisms” consisting of asexually produced bodies (termed zooids) that are homologous to solitary free-living polyps and medusae, but that share a common gastrovascular cavity [@dunn2006evolution;@mackie1963; @mackie1986aggregates;@mackie1987siphonophore;@totton1965synopsis]. The extreme specialization of siphonophore zooids has been of central interest to zoologists since the 19th century, in part because these zooids are highly interdependent [@Beklemishev1969; @mackie1963]. Siphonophore zooids have been likened to animal organs, with each functionally specialized zooid performing distinct roles within the colony [@mackie1963]. While this analogy works well for understanding the function of zooids within the colony as a whole, the analogy falls short in terms of explaining the evolutionary origin and development of these biological units: functionally specialized zooids are evolutionarily homologous to free living organisms, they are multicellular, possess distinct zooid specific cell types, and show regional subfunctionalization [@church2015histology;@Mackie1960;@totton1965synopsis]. While the developmental mechanisms generating zooids are very different in different clades of colonial animals [@carre1967; @carre1969etude; @carre1991complete; @dunn2006evolution; @siebert2015stem], the evolutionary processes acting on zooids may be similar to those acting on other modular biological units such as cell type, tissue, and organ [@hiebert2020coloniality]. The cellular and molecular processes underlying the patterning and molecular function of functionally specialized cnidarian zooids remain an open biological question. In recent years there has been a focus in particular on differential gene expression patterns found in different functionally specialized zooids [@sanders2014differential; @sanders2015interspecific; @Siebert:2011gv].

@@

(ref:zooids-definition) Schematic of the siphonophore *Nanomia bijuga*, with highlighted zooids and pneumatophore, with explanations of known function. NGZ - nectosomal growth zone. SGZ - siphosomal growth zone. Diagram by Freya Goetz (http://commons.wikimedia.org/wiki/File:Nanomia_bijuga_whole_animal_and_growth_zones.svg)

@@

Second, we further account for gene and species effects to compare gene expression across genes and species. Normalized read counts produced by RNA-seq experiments are proportional to gene expression, but they are also impacted by factors that can vary across genes and species such as gene sequence and length. Therefore, to estimate and compare expected gene counts it is necessary to model such factors with unknown species and gene-specific counting-efficiency coefficients [@dunn2013phylogenetic]. A direct comparison of expected gene counts among species can be misleading if differences in counts are simply due to differences in counting efficiency and not due to differences in expression. To address this issue, we use ratios of expected counts to compare expression across genes and species. Using ratios of counts, we are able to eliminate species- and gene- specific counting efficiency within species, as the unknown counting efficiency factor is in both the numerator and the denominator, and is thus removed prior to comparisons among species [@dunn2013phylogenetic].

@@

Because the history of genes is characterized by more complex scenarios involving gene duplication and speciation, we introduce a new method that takes advantage of this rich history to examine the evolution of gene expression across more genes and species. We call this method Species Tree Edge Extraction and Layover (STEEL) (Fig. \@ref(fig:methods-comp)), as we map expression data to gene phylogenies and identify equivalent branches in the species tree which are descended from speciation events (hereafter, species-equivalent branches) in order to make valid comparisons across species. While SOFT is a gene tree filtering approach, where entire trees are discarded due to even a single duplication event, STEEL is a branch filtering approach. This means that STEEL retains many informative branches from trees that would be removed by SOFT, and preserves many more evolutionary comparisons for analyses.

Briefly, the principles of STEEL are as follows: first we identify speciation and duplication events in the gene phylogeny, specifically labelling speciation nodes in the gene tree that correspond to the species tree (Fig. \@ref(fig:methods-comp), step 1). Next, we map gene expression values to the tips of the gene tree (Fig. \@ref(fig:methods-comp), step 2), and use ancestral trait reconstruction methods to reconstruct expression values at the nodes (Fig. \@ref(fig:methods-comp), step 3). Then, we calculate scaled expression values across branches, by subtracting the expression value at the child node from the parent node, and dividing by branch length (branch lengths are calibrated against the species tree) (Fig. \@ref(fig:methods-comp), step 4). Finally, we identify branches in the gene trees that correspond to specific branches in the species tree, that we call 'species-equivalent branches' (Fig. \@ref(fig:methods-comp), step 5). Species-equivalent branches are gene tree branches that have parent and child nodes that are both speciation events and that correspond to speciation events and branches in the species tree. Each branch in the species tree is given a unique letter and the species-equivalent branches in gene trees are given the same letter (Fig. \@ref(fig:methods-comp), step 5). This method enables the selection of specific branches within gene trees that are equivalent to branches within the species tree, and are thus comparable with one another across all gene trees. Unlike the SOFT approach, this approach considers equivalent branches that are descended from speciation events, but that have more complex evolutionary histories. For example, due to deeper gene duplication events, gene trees often contain multiple branches that correspond to the same branch in the species tree (Fig. \@ref(fig:methods-comp), step 5). Our method allows us to consider all of these branches.

Using the methods described above, combined with classical differential gene expression analysis, we conducted differential expression analyses among siphonophore zooids, and one specialized tissue (the pneumatophore, a gas filled float), in seven siphonophore species to address three questions. First, we analyzed differences in gene expression within species to identify patterns that define particular zooid types, and gain insight into the genes that may be playing a role in the structure, maintenance, and functioning of zooids. Second, we examined whether gene expression patterns could distinguish novel, distinct species-specific zooid types within particular species. Third we compared gene expression across the seven siphonophore species, to identify putative clade/lineage or zooid specific expression patterns. 
(ref:methods-comp) Methods used to identify changes in expression across branches in gene trees, called Species Tree Edge Extraction and Layover (STEEL). Step 1, we label each of the nodes in the species phylogeny, and identify equivalent speciation nodes across every gene tree (an exemplar is shown here). Step 2, we map expression values (TPM10K) to the tips . Step 3, we reconstruct ancestral trait expression values at all nodes where expression data are available. Step 4, we calculate scaled change in expression (child node expression - parent node expression / branch length). Branch length is calibrated to the species tree branch lengths. Step 5, we identify branches in gene trees that correspond to equivalent branches in the species tree. There may be more than one branch in a gene tree that corresponds to the same branch in the species tree.

@@

The first component of variation that we assessed was among technical replicates. The technical replicates consist of re-sequenced developing nectophores and developing gastrozooids from the same *Frillagalma vityazi* individual that were spiked in across multiple lanes and runs. Lane and run effects have been proposed as major sources of technical variability in RNA-seq data that may confound observations of biological variation [@auer2010statistical; @mcintyre2011rna]. The differences between technical replicates (Fig. \@ref(fig:technical-replicate)) were found to be much smaller (0.39% variance of expression distance) than the differences between zooids (98.32% variance of expression distance). Differences among technical replicates of the same zooid were correlated with library size and run, not by lane.

@@

Within the four best sampled species, we also identified `r x$short_blast_hit %>% unique() %>% length()` higher abundance putative transcription factors (`r length(Bargmannia_TF)` out of `r y[[2]] %>% sum()` in *Bargmannia elongata*, `r length(Frillagalma_TF)` out of `r y[[4]] %>% sum` identified in *Frillagalma vityazi*, `r length(Agalma_TF)` out of `r y[[1]] %>% sum()` in *Agalma elegans* and `r length(Nanomia_TF)` out of `r y[[5]] %>% sum()` identified  in *Nanomia bijuga*). Many of these transcription factors have higher expression in several zooids regardless of developmental stage (both mature and developing zooids), and a subset were found to have higher expression only in particular zooids and developmental stages (Fig. \@ref(fig:transcription-factors)).

@@

A large number of genes were identified as being significantly upregulated in mature gastrozooids or  tentacular palpons in *Physalia physalis* (`r dge_sig[[6]]$TenpalmatGasmat %>% nrow()` genes in the mature tentacular palpon and `r dge_sig[[6]]$GasmatTenpalmat %>% nrow()` genes in the mature gastrozooid). A number of genes were found to have higher transcript abundance in the mature tentacular palpon relative to all other tissues, of which, `r Physalia_Tenpalmat_unique %>% length()` genes were found to be upregulated relative to other tissues that were not shared with the gastrozooid. In the gastrozooid `r Physalia_Gasmat_unique %>% length()` genes were found to be upregulated relative to other tissues that were not shared with the tentacular palpon. A number of genes upregulated in the tentacular palpon are uncharacterized, however we identified 46 putative toxins in the tentacular palpon, including hemostasis interfering and platelet aggregation activating toxins, phospholipases, serineproteases, hydrolases, metalloendoproteases, calglandulin-like genes, and a neurotoxin. By contrast, in the gastrozooid, we found 59 upregulated putative toxins, including pore-forming Conoporin-Cn1-like and Tereporin-Ca1-like toxins, multiple neurotoxins, hydrolases, serine proteases, toxins likely involved in the promotion of blood coagulation and inhibition of platelet aggregation. Reflecting the role of the gastrozooid in digestion, we also find upregulation of digestive enzymes, including chymotrypsin-like genes.

Between the white mature gastrozooid and the yellow mature gastrozooid in *Bargmannia elongata*, few upregulated genes were identified (`r dge_sig[[2]]$GaswhimatGasyelmat %>% nrow()` genes were up in "white" mature gastrozooids relative to `r dge_sig[[2]]$GasyelmatGaswhimat %>% nrow()` genes in "yellow" gastrozooids). Among genes that were found to be significantly upregulated in either "white" or "yellow" gastrozooids relative to all other tissues, `r Bargmannia_Gasyelmat_unique %>% length()` genes were unique to "yellow" gastrozooids and not found in "white" gastrozooids, and `r Bargmannia_Gaswhimat_unique %>% length()` genes were found in "white" gastrozooids and not found in "yellow" gastrozooids.

Finally, in *Agalma elegans* very few significantly upregulated genes were identified between the B palpon and gastric palpons  (`r dge_sig[[1]]$PalBmatPalmat %>% nrow()` had higher transcript abundance in B palpons and `r dge_sig[[1]]$PalmatPalBmat %>% nrow()` had higher transcript abundance in gastric palpons. All three genes have no significant blast hit and did not map to any gene trees). Genes were identified that are upregulated in B palpons relative to all other zooids (gastric palpons were excluded). Of this, `r Agalma_PalponB_unique %>% length()` genes were found to be differentially expressed in the B palpon. Most of these genes overlapped with those found to be differentially expressed in the gastric palpon relative to all other zooids (`r intersect(Agalma_PalponB_unique,Agalma_Palpon_unique) %>% length()` genes).

@@

(ref:SOFT-plot) Results of Strict Ortholog Filtering of Trees analyses. Top row: results using TPM10K expression values; Bottom row: results using TPM10K expression ratios. For both analyses of expression variance explained by species and tissue, we identified highly species-variable genes (SVG, red) and highly tissue-variable genes (TVG, blue). In SVGs, the proportion of variance explained by species is two times greater than that explained by tissues; while in TVGs the proportion of variance explained by tissues is two times greater than that explained by species. All SVGs and TVGs are genes for which both species and tissue explain greater than 75% of variance. A. PCA plot of expression values (TPM10k) of `r res %>% nrow()` strict ortholog genes, showing a species dominated clustering pattern. B.  The proportion of expression variance explained by species (x-axis) versus the proportion of expression variance explained by tissues (y-axis), using TPM10K expression values. C. PCA plot of expression ratios of `r res2 %>% nrow()` strict ortholog genes, showing a tissue dominated clustering pattern.  D. The proportion of expression variance explained by species (x-axis) versus the proportion of expression variance explained by tissues (y-axis), using expression ratios. Abbreviations: Gasdev = developing gastrozooid, Necdev = developing nectophore, Palmat = mature palpon, Pne = pneumatophore, Gasmat = mature gastrozooid.

@@

Most comparative studies of gene expression focus exclusively on strict 1:1 orthologs, requiring an assessment of orthology across all species. We call this type of analysis Strict Ortholog Filtering of Trees (SOFT), as it limits analyses to a subset of genes with very specific evolutionary histories. Following this approach, we used Orthofinder 2 (v2.4.0) [@emms2019orthofinder] to identify a set of `r Orthologs2$Orthogroup %>% unique() %>% length()` strict orthologs for four of the best sampled species (*Agalma elegans*, *Bargmannia elongata*, *Frillagalma vityazi*, and *Nanomia bijuga*), of which `r stvariance %>% nrow()` orthologs had expression values in four tissues: gastrozooids (developing and mature), nectophores (developing), palpons (mature), and the pneumatophore. Expression values were compared across all of the species. We found that using TPM10K expression values the average behaviour of all orthologs favored a species-dominated clustering pattern (Fig. \@ref(fig:SOFT-plot)A). Following Breschi et al. @breschi2016gene (see methods), we used linear models to identify the proportion of variance that could be explained by species or tissue (Fig. \@ref(fig:SOFT-plot)B), finding out of `r stvariance  %>% nrow()` orthologs, `r stvariance %>% filter(class=="SVG") %>% nrow()` species-variable genes (SVG), and `r stvariance %>% filter(class=="TVG") %>% nrow()` tissue-variable genes (TVG).

The strong species-dominated clustering observed when comparing expression values directly may be due to differences in counting efficiency between species, especially as reference transcriptomes were used [@dunn2013phylogenetic]. To control for this, we used ratios of expression values, using TMP10K expression values of the most commonly sampled zooid, mature gastrozooid, as the denominator [@dunn2013phylogenetic]. Using expression ratios, the average behaviour of all orthologs favored more tissue-dominated clustering pattern (Fig. \@ref(fig:SOFT-plot)C). Using linear models, we identified, out of `r stvariance2  %>% nrow()` orthologs, `r stvariance2 %>% filter(class=="SVG") %>% nrow()` species-variable genes (SVG), and `r stvariance2 %>% filter(class=="TVG") %>% nrow()` tissue-variable genes (TVG) (Fig. \@ref(fig:SOFT-plot)D). GO terms found to be enriched among TVG include embryonic morphogenesis, embryo development, cartilage development, and a number of metabolic and biosynthetic processes. Meanwhile, SVG were enriched in GO terms such as cellular response to stress, DNA repair and cell cycle processes. A list of tissue variable and species variable genes can be found in supplementary files 4 and 5 respectively.

@@

For our Species Tree Edge Extraction and Layover (STEEL) analyses (see Fig. \@ref(fig:methods-comp) and methods), we used transcriptome and genome data from `r species_tree$tip.label %>% length()` cnidarian species to generate a total of `r gene_trees %>% length()` gene trees, of which `r gene_trees_ace_calibrated[!is.na(gene_trees_ace_calibrated)] %>% length()` gene trees passed filtering criteria (see Methods). We used Orthofinder 2 [@emms2019orthofinder] to reconcile the species tree and the gene trees and annotate each gene tree node as either a speciation or duplication events. The number of genes represented in these gene trees is shown in Table \@ref(tab:summarize-libraries). The internal nodes on these gene trees consist of `r nodes_ace %>% filter(is.na(species) & Ev=="S") %>% nrow()` speciation events and `r nodes_ace %>% filter(is.na(species) & Ev=="D") %>% nrow()` duplication events.

Expression values were mapped to the tips of the gene trees, and ancestral values were inferred at internal nodes with maximum likelihood. Changes in expression across branches were calculated based on the difference between parent and child nodes, scaled by the branch length [@dunn2018pairwise]. We focused on expression in a subset of zooids and tissues that are common across the sampled species: gastrozooids (developing and mature), nectophores (developing), palpons (mature), and the pneumatophore. The distribution of expression changes along species-equivalent branches are shown in figure \@ref(fig:changes-branches). To examine if expression values in our dataset evolve under Brownian Motion (BM), we simulated data set of random expression values with empirically derived mean and standard-deviation values for each gene tree. Under this model, expression variance accumulates among linages on the gene tree as a function of time, and is used as a model of evolution under drift, as well as some forms of natural selection [@felsenstein1973maximum;@revell2008testing;@revell2008phylogenetic].

>>>>

**Background**

Siphonophores are highly complex colonial organisms, consisting of asexually-produced bodies (called zooids) that share a common gastrovascular cavity but that are functionally specialized to specific tasks, including feeding, swimming, and reproducing. Though this extreme functional specialization has captivated biologists for generations, the genomic underpinnings of this phenomenon remain poorly understood. We use RNA-seq to investigate gene expression patterns in five zooids, and one specialized tissue (the pneumatophore, a gas filled float), across seven siphonophore species. To overcome the analytical challenges of studying the evolution of gene expression across genes and species, we introduce a novel expression metric and two new analytical approaches.

**Results**

Within species, we identified hundreds of zooid-specific genes, as well as a number of putative transcription factors showing higher abundance in particular zooids and developmental stages. In species with unique zooid types, we found significant differences in expression profiles between zooids with different function (i.e., feeding vs. tentacle in Physalia physalis), but no differences between distinct zooids with similar functions (e.g., feeding zooids in Bargmannia elongata and Agalma elegans). Across all species, we found that gene expression patterns tended to be largely consistent among species rather than zooids, and found evidence of large lineage specific shifts in gene expression.

**Conclusions**

Our findings show that there are distinct expression patterns defining each functionally specialized zooid or tissue. In species-specific instances of zooid diversification, we could distinguish novel zooids with distinct morphological and functional differences, but were unable to gain insights into differences between zooids with small differences in morphology or colony placement. Using comparative approaches, we also identified shared zooid-specific gene expression patterns across species, and within a given zooid type, identified instances of clade/species-specific expression patterns.

@@

Colonial animals, consisting of genetically identical bodies that are physically and physiologically connected, can be found across the metazoan tree of life [@hiebert2020coloniality]. Functional specialization of such bodies has evolved multiple times within colonial animals, with siphonophores in particular showing the greatest diversity of functionally specialized bodies [@Beklemishev1969]. Siphonophores are highly complex, colonial “superorganisms” consisting of asexually produced bodies (termed zooids) that are homologous to solitary free-living polyps and medusae (the typical body forms in Cnidaria), but that share a common gastrovascular cavity [@dunn2006evolution;@mackie1963; @mackie1986aggregates;@mackie1987siphonophore;@totton1965synopsis]. The extreme specialization of siphonophore zooids has been of central interest to zoologists since the 19th century, in part because these zooids are highly interdependent [@Beklemishev1969; @mackie1963]. Siphonophore zooids have been likened to animal organs, with each functionally specialized zooid performing distinct roles within the colony [@mackie1963]. While this analogy works well for understanding the function of zooids within the colony as a whole, the analogy falls short in terms of explaining the evolutionary origin and development of these biological units: functionally specialized zooids are evolutionarily homologous to free living organisms, they are multicellular, possess distinct zooid specific cell types, and show regional subfunctionalization [@church2015histology;@Mackie1960;@totton1965synopsis]. While the developmental mechanisms generating zooids are very different in different clades of colonial animals [@carre1967; @carre1969etude; @carre1991complete; @dunn2006evolution; @siebert2015stem], the evolutionary processes acting on zooids may be similar to those acting on other modular biological units such as cell type, tissue, and organ [@hiebert2020coloniality]. The cellular and molecular processes underlying the patterning and molecular function of functionally specialized cnidarian zooids remain an open biological question. In recent years there has been a focus in particular on differential gene expression patterns found in different functionally specialized zooids [@sanders2014differential; @sanders2015interspecific; @Siebert:2011gv].

@@

(ref:zooids-definition) Schematic of a siphonophore colony, zooid organization, and zooid function. Here, we illustrate the siphonophore *Nanomia bijuga*, with highlighted zooids and pneumatophore, and explanations of known function. NGZ - nectosomal growth zone. SGZ - siphosomal growth zone. Diagram by Freya Goetz (http://commons.wikimedia.org/wiki/File:Nanomia_bijuga_whole_animal_and_growth_zones.svg)

@@

Second, we further account for gene and species effects to compare gene expression across genes and species. Normalized read counts produced by RNA-seq experiments are proportional to gene expression, but they are also impacted by factors that can vary across genes and species such as gene sequence and length. Therefore, to estimate and compare expected gene counts it is necessary to model such factors with unknown species and gene-specific counting-efficiency coefficients [@dunn2013phylogenetic]. A direct comparison of expected gene counts among species, however, can be misleading if differences in counts are simply due to differences in counting efficiency and not due to differences in expression. To address this issue, we use ratios of expected counts to compare expression across genes and species. Using ratios of counts, we are able to eliminate species- and gene- specific counting efficiency within species, as the unknown counting efficiency factor is in both the numerator and the denominator, and is thus removed prior to comparisons among species [@dunn2013phylogenetic].

@@

Because the history of genes is characterized by more complex scenarios involving gene duplication and speciation, we introduce a new method that takes advantage of this rich history to examine the evolution of gene expression across more genes and species. We call this method Species Tree Edge Extraction and Layover (STEEL) (Fig. \@ref(fig:methods-comp)), as we map expression data to gene phylogenies and identify equivalent branches in the species tree which are descended from speciation events in order to make valid comparisons across species. While SOFT is a gene tree filtering approach, where entire trees are discarded due to even a single duplication event, STEEL is a branch filtering approach. This means that STEEL retains many informative branches from trees that would be removed by SOFT, and preserves many more evolutionary comparisons for analyses.

Briefly, the principles of STEEL are as follows: first we identify speciation and duplication events in a given gene tree, specifically labelling nodes that correspond to speciation events in the species tree (Fig. \@ref(fig:methods-comp), step 1). Next, we map gene expression values to the tips of the gene tree (Fig. \@ref(fig:methods-comp), step 2), and use phylogenetic methods to reconstruct expression values at the internal nodes (Fig. \@ref(fig:methods-comp), step 3). Expression values are mapped and reconstructed for each zooid/tissue separately, with the assumption that the structure is homologous across species. Then, we calculate scaled expression values across branches, by subtracting the expression value at the child node from the parent node, and dividing by branch length (branch lengths are calibrated against branches in the species tree) (Fig. \@ref(fig:methods-comp), step 4). Finally, we identify branches in the gene trees that correspond to equivalent branches in the species tree (hereafter species-equivalent branches) (Fig. \@ref(fig:methods-comp), step 5). Species-equivalent branches are gene tree branches with parent and child nodes that are both speciation events and that correspond to branches in the species tree. Each branch in the species tree is given a unique identifier (i.e., a letter) and the species-equivalent branches in gene trees are given the same identifier (Fig. \@ref(fig:methods-comp), step 5). This method enables the selection of branches within gene trees that are equivalent to branches within the species tree, and are thus comparable with one another across all gene trees. Unlike the SOFT approach, this approach considers equivalent branches that are descended from speciation events, but that have more complex evolutionary histories. For example, due to deeper gene duplication events, gene trees often contain multiple branches that correspond to the same branch in the species tree (Fig. \@ref(fig:methods-comp), step 5). Our method allows us to consider all of these branches.

Using the methods described above, combined with classical differential gene expression analysis, we conducted differential expression analyses among siphonophore zooids, and one specialized tissue (the pneumatophore, a gas filled float), in seven siphonophore species to address three questions. First, we analyzed differences in gene expression within species to identify patterns that define particular zooid types, and gain insight into the genes that may be playing a role in the structure, maintenance, and functioning of zooids. Second, we examined whether gene expression patterns could distinguish novel, distinct species-specific zooid types within particular species. Third we compared gene expression across the seven siphonophore species, to identify putative clade/lineage or zooid specific expression patterns.

(ref:methods-comp) New phylogenetic approach to identify evolutionary changes in gene expression, called Species Tree Edge Extraction and Layover (STEEL). Step 1, we label each of the nodes in the species tree, and identify equivalent speciation nodes across every gene tree (an exemplar is shown here). Step 2, we map expression values (TPM10K) to the tips (expression values are mapped and reconstructed for each homologous zooid separately). Step 3, we reconstruct ancestral trait expression values at all internal nodes where expression data are available. Step 4, we calculate scaled change in gene expression (child node expression - parent node expression / branch length). Branch length is calibrated to the species tree branch lengths. Step 5, we identify branches in gene trees that correspond to equivalent branches in the species tree. There may be more than one branch in a gene tree that corresponds to the same branch in the species tree.

@@

The first component of variation that we assessed was among technical replicates. The technical replicates consist of re-sequenced developing nectophores and developing gastrozooids from the same *Frillagalma vityazi* individual that were spiked in across multiple lanes and runs. Lane and run effects have been proposed as major sources of technical variability in RNA-seq data that may confound observations of biological variation [@auer2010statistical; @mcintyre2011rna]. The differences between technical replicates (Fig. \@ref(fig:technical-replicate)) were much smaller (0.39% variance of expression distance) than the differences between zooids (98.32% variance of expression distance). Differences among technical replicates of the same zooid were correlated with library size and run, not by lane.

@@

Within the four best sampled species, we also identified `r x$short_blast_hit %>% unique() %>% length()` higher abundance putative transcription factors (`r length(Bargmannia_TF)` out of `r y[[2]] %>% sum()` in *Bargmannia elongata*, `r length(Frillagalma_TF)` out of `r y[[4]] %>% sum` identified in *Frillagalma vityazi*, `r length(Agalma_TF)` out of `r y[[1]] %>% sum()` in *Agalma elegans* and `r length(Nanomia_TF)` out of `r y[[5]] %>% sum()` identified  in *Nanomia bijuga*). Many of these transcription factors have higher expression in several zooids regardless of developmental stage (both mature and developing zooids), and a subset have higher expression only in particular zooids and developmental stages (Fig. \@ref(fig:transcription-factors)).

@@

In *Physalia physalis* `r dge_sig[[6]]$TenpalmatGasmat %>% nrow()` genes showed higher transcript abundance in the mature tentacular palpon, compared to `r dge_sig[[6]]$GasmatTenpalmat %>% nrow()` genes in the mature gastrozooid. A number of genes had higher transcript abundance in the mature tentacular palpon relative to all other tissues, of which, `r Physalia_Tenpalmat_unique %>% length()` genes were upregulated relative to other tissues that were not shared with the gastrozooid. In the gastrozooid `r Physalia_Gasmat_unique %>% length()` genes were upregulated relative to other tissues that were not shared with the tentacular palpon. A number of genes upregulated in the tentacular palpon are uncharacterized, however we identified 46 putative toxins in the tentacular palpon, including hemostasis interfering and platelet aggregation activating toxins, phospholipases, serineproteases, hydrolases, metalloendoproteases, calglandulin-like genes, and a neurotoxin. By contrast, in the gastrozooid, we found 59 upregulated putative toxins, including pore-forming Conoporin-Cn1-like and Tereporin-Ca1-like toxins, multiple neurotoxins, hydrolases, serine proteases, toxins likely involved in the promotion of blood coagulation and inhibition of platelet aggregation. Reflecting the role of the gastrozooid in digestion, we also find upregulation of digestive enzymes, including chymotrypsin-like genes.

Between the white mature gastrozooid and the yellow mature gastrozooid in *Bargmannia elongata*, few upregulated genes were identified (`r dge_sig[[2]]$GaswhimatGasyelmat %>% nrow()` genes were up in "white" mature gastrozooids relative to `r dge_sig[[2]]$GasyelmatGaswhimat %>% nrow()` genes in "yellow" gastrozooids). Among genes that were significantly upregulated in either "white" or "yellow" gastrozooids relative to all other tissues, `r Bargmannia_Gasyelmat_unique %>% length()` genes were unique to "yellow" gastrozooids and not found in "white" gastrozooids, and `r Bargmannia_Gaswhimat_unique %>% length()` genes were found in "white" gastrozooids and not found in "yellow" gastrozooids.

Finally, in *Agalma elegans* very few significantly upregulated genes were identified between the B palpon and gastric palpons  (`r dge_sig[[1]]$PalBmatPalmat %>% nrow()` had higher transcript abundance in B palpons and `r dge_sig[[1]]$PalmatPalBmat %>% nrow()` had higher transcript abundance in gastric palpons. All three genes have no significant blast hit and did not map to any gene trees). Genes were identified that are upregulated in B palpons relative to all other zooids (gastric palpons were excluded). Of this, `r Agalma_PalponB_unique %>% length()` genes were differentially expressed in the B palpons. Most of these genes overlapped with those differentially expressed in the gastric palpons relative to all other zooids (`r intersect(Agalma_PalponB_unique,Agalma_Palpon_unique) %>% length()` genes).

@@

(ref:SOFT-plot) Results of Strict Ortholog Filtering of Trees analyses. Top row: results using TPM10K expression values; Bottom row: results using TPM10K expression ratios. For both analyses of expression variance explained by species and tissue (right-hand column), we identified highly species-variable genes (SVG, red) and highly tissue-variable genes (TVG, blue). In SVGs, the proportion of variance explained by species is two times greater than that explained by tissues; while in TVGs the proportion of variance explained by tissues is two times greater than that explained by species. All SVGs and TVGs are genes for which both species and tissue explain greater than 75% of variance. A. PCA plot of expression values (TPM10K) of `r res %>% nrow()` strict ortholog genes, showing a species dominated clustering pattern. B.  The proportion of expression variance explained by species (x-axis) versus the proportion of expression variance explained by tissues (y-axis), using TPM10K expression values. C. PCA plot of expression ratios of `r res2 %>% nrow()` strict ortholog genes, showing a tissue dominated clustering pattern.  D. The proportion of expression variance explained by species (x-axis) versus the proportion of expression variance explained by tissues (y-axis), using expression ratios. Abbreviations: Gasdev = developing gastrozooid, Necdev = developing nectophore, Palmat = mature palpon, Pne = pneumatophore, Gasmat = mature gastrozooid.

@@

Most comparative studies of gene expression focus exclusively on strict 1:1 orthologs, requiring an assessment of orthology across all species. We call this type of analysis Strict Ortholog Filtering of Trees (SOFT), as it limits analyses to a subset of genes with very specific evolutionary histories [e.g @Brawand:2011du; @cardoso2019gene; @Levin:2016bp; @yang2013organ]. Following this approach, we used Orthofinder 2 (v2.4.0) [@emms2019orthofinder] to identify a set of `r Orthologs2$Orthogroup %>% unique() %>% length()` strict orthologs for four of the best sampled species (*Agalma elegans*, *Bargmannia elongata*, *Frillagalma vityazi*, and *Nanomia bijuga*), of which `r stvariance %>% nrow()` orthologs had expression values in four zooids/tissues: gastrozooids (developing and mature), nectophores (developing), palpons (mature), and the pneumatophore. Using TPM10K alone, we found that orthologs clustered largely by species rather than zooid/ tissue (Fig. \@ref(fig:SOFT-plot)A). Following Breschi et al. @breschi2016gene (see methods), we used linear models to identify the proportion of variance that could be explained by species or zooid/tissue (Fig. \@ref(fig:SOFT-plot)B), finding out of `r stvariance  %>% nrow()` orthologs, `r stvariance %>% filter(class=="SVG") %>% nrow()` species-variable genes (SVG), and `r stvariance %>% filter(class=="TVG") %>% nrow()` zooid/tissue-variable genes (TVG).

The strong species-dominated clustering observed when comparing expression values directly may be due to differences in counting efficiency between species, especially as we used reference transcriptomes [@dunn2013phylogenetic]. To account for this, we used ratios of expression values, using TMP10K expression values of the most commonly sampled zooid -- mature gastrozooid -- as the denominator [@dunn2013phylogenetic]. Using this approach, we found that orthologs clustered largely by zooid/tissue rather than species (Fig. \@ref(fig:SOFT-plot)C). Using linear models, we identified, out of `r stvariance2  %>% nrow()` orthologs, `r stvariance2 %>% filter(class=="SVG") %>% nrow()` species-variable genes (SVG), and `r stvariance2 %>% filter(class=="TVG") %>% nrow()` tissue-variable genes (TVG) (Fig. \@ref(fig:SOFT-plot)D). GO terms enriched among TVG included embryonic morphogenesis, embryo development, cartilage development, and a number of metabolic and biosynthetic processes. Meanwhile, SVG were enriched in GO terms such as cellular response to stress, DNA repair and cell cycle processes. A list of TVG and SVG can be found in supplementary files 4 and 5 respectively.

@@

For our Species Tree Edge Extraction and Layover (STEEL) analyses (see Fig. \@ref(fig:methods-comp) and methods), we used transcriptome and genome data from `r species_tree$tip.label %>% length()` cnidarian species to generate a total of `r gene_trees %>% length()` gene trees, of which `r gene_trees_ace_calibrated[!is.na(gene_trees_ace_calibrated)] %>% length()` gene trees passed filtering criteria (see Methods). We used Orthofinder 2 [@emms2019orthofinder] to reconcile the species tree and the gene trees and annotate each gene tree node as either a speciation or duplication events. The number of genes represented in these gene trees is shown in Table \@ref(tab:summarize-libraries). The internal nodes on these gene trees consist of `r nodes_ace %>% filter(is.na(species) & Ev=="S") %>% nrow()` speciation events and `r nodes_ace %>% filter(is.na(species) & Ev=="D") %>% nrow()` duplication events. Expression values (TPM10K) were mapped to the tips of the gene trees for each zooid/tissue separately, and ancestral values were inferred at internal nodes with maximum likelihood. We focused on expression in a subset of zooids and tissues that are common across the sampled species: gastrozooids (developing and mature), nectophores (developing), palpons (mature), and the pneumatophore. The distribution of expression changes along species-equivalent branches are shown in figure \@ref(fig:changes-branches).
